Qualitative Probabilistic Programming

### 2015-06-05 paper/paper.tex
finishing up

  In the following, we first specify the syntax used to write Quipp programs,
  including the notation for unknown variables and functions.
  We describe the class of exponential family variables and functions that our system can learn,
  and present the expectation maximization algorithm used to learn them.
  We then demonstrate the expressiveness of our language, and the broad
  applicability of our algorithm, by writing some of the most common machine learning models
  in Quipp: clustering, naive Bayes, factor analysis, a Hidden Markov model, Latent Dirichlet Allocation, and
  a neural net.

@@

  To demonstrate, let us run this example on the Iris dataset, which consists of 150 points \cite{Lichman:2013}.  When we run the program on this data, we infer the parameters to the random function \texttt{getPoint}.
  In this case, \texttt{getPoint} is a linear function with Gaussian noise, so it will naturally
  split the data into 2 clusters with equal variance, as seen in Figure 1.

  \begin{figure}[h]
  \begin{center}
    \includegraphics[scale=0.5]{../plots/irisclusters_orig.png}
  \end{center}
  \caption{Clustering the Iris dataset}
\end{figure}

@@

    Additionally, we provide a $\texttt{randomValue}$ function for convenience, which
    returns a sample from the ``default'' distribution for a given type.  Table 1 shows the types
    in Quipp and their corresponding sufficient statistics, features, and English descriptions.

@@

  \section{Example: improving the clustering model}
  Using this knowledge about the form of the random functions, we can improve the clustering model from before.
  As observed in Figure 1, the two clusters are forced to have the
  same variance. They do not fit the data well, since the data has a different
  shape in each location.  To fix this problem, we can substitute the following
  model, which uses a separate random function (and therefore a separate axis-aligned Gaussian distribution) for each cluster:
  \begin{lstlisting}
var Cluster = Categorical(2);
var Point = Vector(2, Double);
var getPointFunctions = [randFunction(Point), randFunction(Point)];

var model = function() {
  var cluster = randomValue(Cluster);
  observe(getPointFunctions[cluster]);
};
\end{lstlisting}
  \begin{figure}[h]
  \begin{center}
    \includegraphics[scale=0.5]{../plots/irisclusters_indep.png}
  \end{center}
  \caption{Clustering the Iris dataset with independent variances}
\end{figure}
  Using this model, we get the clusters shown in Figure 2.
  This model (with the parameters filled in) is estimated to assign probability density $e^{-337}$ to the data, yielding a perplexity of 0.1058.  This is a significantly improved fit.

@@

  Here, there are 2 random functions for each cluster, one to get the $x$ coordinate and one to get the $y$ coordinate (whose distribution may depend linearly on the $x$ coordinate).  This allows $x$ and $y$ to be correlated, improving fit, as seen in Figure 3.
  This model is estimated to assign probability $e^{-269}$ to the data, yielding a perplexity of 0.166.  This is a large improvement from the previous model.

  \begin{figure}[h]
  \begin{center}
    \includegraphics[scale=0.5]{../plots/irisclusters_dep.png}
  \end{center}
  \caption{Clustering the Iris dataset with multivariate Gaussian distributions}
\end{figure}

@@

  \subsection{Clustering}

@@

In this example, we cluster 2d points into 3 different clusters.  Given a cluster, the distribution for a point is some independent Gaussian distribution.  This is similar to fuzzy c-means clustering.

@@

\subsection{Naive Bayes}

@@

  \subsection{Factor analysis}
{\small
\begin{lstlisting}
var Factors = Vector(2, Double);
var Point = Vector(5, Double);
var getPoint = randFunction(Factors, Point);

@@

The factor analysis model is very similar to the clustering model.  The main difference is that we replace the categorical \texttt{ClusterType} type with a vector type.  \texttt{randomValue(Factors)} will return a vector from the standard multivariate normal distribution.  This results in the model attempting to predict each point as an affine function of a vector of standard normal values, with Gaussian noise.

@@

  \subsection{Hidden Markov model}
{\small
\begin{lstlisting}
var chainLength = 20;
var State = Categorical(2);
var Obs = Categorical(4);
var transFun = randFunction(State, State);
var obsFun = randFunction(State, Obs);

@@

In this example, we use the unknown function \texttt{transFun} for state transitions and \texttt{obsFun} for observations.  This means that we will learn both the state transition matrix and the observation matrix.

@@

\subsection{Latent Dirichlet allocation}
{\small
\begin{lstlisting}
var maxWordsPerDocument = 100;
var Topic = Categorical(3);
var Word = Categorical(10);
var topicToWord = randFunction(Topic, Word);

@@

For latent Dirichlet allocation, we use the unknown function \texttt{topicToWord} to map latent topics to word distributions.  We will learn a different categorical distribution for each topic.

  \begin{figure}[h]
\begin{center}
  \includegraphics[scale=0.5]{../plots/accuracy_lda.png}
\end{center}
\caption{Regret-per-sample quantiles for latent Dirichlet allocation}
\end{figure}
\subsection{Neural network}

{\small
\begin{lstlisting}
var Input = Vector(30, Bool);
var Hidden = Vector(10, Bool);
var Output = Bool;

var inputToHidden = randFunction(Input, Hidden);
var hiddenToOutput = randFunction(Hidden, Output);

var model = function() {
  var inputLayer = observe(randomValue, Input);
  var hiddenLayer = inputToHidden(inputs[sampIndex]);
  observe(hiddenToOutput, hiddenLayer);
};
\end{lstlisting}
}

>>>>

  In the following, we first specify the syntax used to write Quipp programs,
  including the notation for unknown variables and functions.
  We describe the class of exponential family variables and functions that our system can learn,
  and present the expectation maximization algorithm used to learn them.
  We then demonstrate the expressiveness of our language, and the broad
  applicability of our algorithm, by writing some of the most common machine learning models
  in Quipp: clustering, naive Bayes, factor analysis, a Hidden Markov model and Latent Dirichlet Allocation.

@@

  To demonstrate, let us run this example on the Iris dataset, which consists of 150 points \cite{Lichman:2013}.  When we run the program on this data, we infer the parameters to the random function \texttt{getPoint}.
  In this case, \texttt{getPoint} is a linear function with Gaussian noise, so it will naturally
  split the data into 2 clusters with equal variance, as seen in Figure \ref{iris-orig}.

  \begin{figure}[h]
  \begin{center}
    \includegraphics[scale=0.5]{../plots/irisclusters_orig.png}
  \end{center}
  \caption{Clustering the Iris dataset}
  \label{iris-orig}
\end{figure}

@@

    Additionally, we provide a $\texttt{randomValue}$ function for convenience, which
    returns a sample from the ``default'' distribution for a given type.  Table \ref{type-table} shows the types
    in Quipp and their corresponding sufficient statistics, features, and English descriptions.

@@

  \section{Example: improving the clustering model}
  Using this knowledge about the form of the random functions, we can improve the clustering model from before.
  As observed in Figure \ref{iris-orig}, the two clusters are forced to have the
  same variance. They do not fit the data well, since the data has a different
  shape in each location.  To fix this problem, we can substitute the following
  model, which uses a separate random function (and therefore a separate axis-aligned Gaussian distribution) for each cluster:
  \begin{lstlisting}
var Cluster = Categorical(2);
var Point = Vector(2, Double);
var getPointFunctions = [randFunction(Point), randFunction(Point)];

var model = function() {
  var cluster = randomValue(Cluster);
  observe(getPointFunctions[cluster]);
};
\end{lstlisting}
  \begin{figure}[h]
  \begin{center}
    \includegraphics[scale=0.5]{../plots/irisclusters_indep.png}
  \end{center}
  \caption{Clustering the Iris dataset with independent variances}
  \label{iris-indep}
\end{figure}
Using this model, we get the clusters shown in Figure \ref{iris-indep}.
  This model (with the parameters filled in) is estimated to assign probability density $e^{-337}$ to the data, yielding a perplexity of 0.1058.  This is a significantly improved fit.

@@

Here, there are 2 random functions for each cluster, one to get the $x$ coordinate and one to get the $y$ coordinate (whose distribution may depend linearly on the $x$ coordinate).  This allows $x$ and $y$ to be correlated, improving fit, as seen in Figure \label{iris-dep}.
  This model is estimated to assign probability $e^{-269}$ to the data, yielding a perplexity of 0.166.  This is a large improvement from the previous model.

  \begin{figure}[h]
  \begin{center}
    \includegraphics[scale=0.5]{../plots/irisclusters_dep.png}
  \end{center}
  \caption{Clustering the Iris dataset with multivariate Gaussian distributions}
  \label{iris-dep}
\end{figure}

@@

\FloatBarrier
  \subsection{Clustering}

@@

In this example, we cluster 2d points into 3 different clusters.  Given a cluster, the distribution for a point is some independent Gaussian distribution.  This is similar to fuzzy c-means clustering.  Regret usually reduces quite a lot as iterations increase.

@@

\FloatBarrier
\subsection{Naive Bayes}

@@

\FloatBarrier
  \subsection{Factor analysis}
{\small
\begin{lstlisting}
var Factors = Vector(2, Double);
var Point = Vector(5, Double);
var getPoint = randFunction(Factors, Point);

@@

The factor analysis model is very similar to the clustering model.  The main difference is that we replace the categorical \texttt{ClusterType} type with a vector type.  \texttt{randomValue(Factors)} will return a vector from the standard multivariate normal distribution.  This results in the model attempting to predict each point as an affine function of a vector of standard normal values, with Gaussian noise.  Regret reduces as iterations increase.

@@

\FloatBarrier
  \subsection{Hidden Markov model}
{\small
\begin{lstlisting}
var chainLength = 20;
var State = Categorical(2);
var Obs = Categorical(4);
var transFun = randFunction(State, State);
var obsFun = randFunction(State, Obs);

@@

In this example, we use the unknown function \texttt{transFun} for state transitions and \texttt{obsFun} for observations.  This means that we will learn both the state transition matrix and the observation matrix.  Regret starts low and does not change over time, indicating that only one EM iteration was necessary to accurately estimate the parameters.

@@

\FloatBarrier
\subsection{Latent Dirichlet allocation}
{\small
\begin{lstlisting}
var maxWordsPerDocument = 100;
var Topic = Categorical(3);
var Word = Categorical(10);
var topicToWord = randFunction(Topic, Word);

@@

For latent Dirichlet allocation, we use the unknown function \texttt{topicToWord} to map latent topics to word distributions.  We will learn a different categorical distribution for each topic.  Regret generally reduces over iterations.

  \begin{figure}[h]
\begin{center}
  \includegraphics[scale=0.5]{../plots/accuracy_lda.png}
\end{center}
\caption{Regret-per-sample quantiles for latent Dirichlet allocation}
\end{figure}

% \FloatBarrier
% \subsection{Neural network}
% 
% {\small
% \begin{lstlisting}
% var Input = Vector(30, Bool);
% var Hidden = Vector(10, Bool);
% var Output = Bool;
% 
% var inputToHidden = randFunction(Input, Hidden);
% var hiddenToOutput = randFunction(Hidden, Output);
% 
% var model = function() {
%   var inputLayer = observe(randomValue, Input);
%   var hiddenLayer = inputToHidden(inputs[sampIndex]);
%   observe(hiddenToOutput, hiddenLayer);
% };
% \end{lstlisting}
% }
